---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pcpr

<!-- badges: start -->
<!-- badges: end -->

The R package `pcpr` implements Principal Component Pursuit (PCP), a robust
dimensionality reduction technique, for pattern recognition tailored to
environmental health data. The statistical methodology and computational details
are provided in [Gibson et al. (2022)](https://doi.org/10.1289/EHP10479).

## Installation

You can install the development version of pcpr from GitHub with:

``` r
# install.packages("pak")
pak::pak("Columbia-PRIME/pcpr")
```

## Modeling overview

PCP algorithms model an observed exposure matrix $D$ as the sum of three underlying
ground-truth matrices:

$$\underset{\text{mixture}}{D_{n \times p}} = \underset{\text{low-rank}}{L_0} + \underset{\text{sparse}}{S_0} + \underset{\text{noise}}{Z_0}$$

a low-rank matrix $L_0$ encoding consistent patterns of exposure, a sparse
matrix $S_0$ isolating unique or outlying exposure events (that cannot be
explained by the consistent exposure patterns), and dense noise $Z_0$. All of
these matrices are of dimension $n \times p$, where $n$ is the number of
observations (e.g. study participants or measurement dates) and $p$ is the
number of exposures (e.g. chemical and/or non-chemical stressors). Beyond this
mixtures model, the main assumption made by PCP is that
$Z_0 \sim \mathcal{N}(\mu, \sigma^2)$ consists of i.i.d. Gaussian noise
corrupting each entry of the overall exposure matrix $D$.

The models in `pcpr` seek to decompose an observed data matrix `D` into estimated
low-rank and sparse components `L` and `S` for use in downstream environmental health
analyses. The functions in `pcpr` are outfitted with three environmental health
(EH)-specific extensions making `pcpr` particularly powerful for EH research:

1. Missing value functionality
2. Leveraging potential limit of detection (LOD) information
3. Non-negativity constraint on the estimated `L` matrix

## Usage

`pcpr` supplies all of the functions needed to incorporate PCP into an
environmental health analyses. Let's work through a short simulated example
first.

The `sim_data()` function lets users generate simple mixtures models for
quick experimentation. Let's use it to simulate a noisy environmental mixture
comprised of 100 observations of 10 chemicals, with three underlying chemical
exposure patterns (or a rank $r = 3$), extreme outlying exposure events along the diagonal of the
matrix, and dense Gaussian noise corrupting all the measurements in the matrix:

```{r sim data}
library(pcpr)

data <- sim_data(n = 100, p = 10, r = 3, sparse_nonzero_idxs = seq(1, 1000, 101), sigma = 0.05)
D <- data$D
L_0 <- data$L
S_0 <- data$S
Z_0 <- data$Z
```

We can see `L_0` has 3 underlying patterns using the `matrix_rank()` function:

```{r matrix rank}
matrix_rank(L_0)
```

Because our mixtures data is often incomplete in practice, let's simulate a random 5% of the values as missing (`NA`) with the `corrupt_mat_randomly()` function:

```{r corrupt mat randomly}
corrupted_data <- corrupt_mat_randomly(D, perc = 0.05)
D_tilde <- corrupted_data$D_tilde
D_tilde_0fill <- D_tilde
D_tilde_0fill[is.na(D_tilde)] <- 0
```

There are two PCP algorithms shipped with `pcpr`: the convex `root_pcp()`
[[Zhang et al. (2021)](https://proceedings.neurips.cc/paper/2021/hash/f65854da4622c1f1ad4ffeb361d7703c-Abstract.html)]
and non-convex `rrmc()` [[Cherapanamjeri et al. (2017)](https://proceedings.mlr.press/v70/cherapanamjeri17a.html)].
To figure out which model would be best for our data, let's inspect the singular values of our
observed mixture using the `sing()` method:

```{r sing}
plot(sing(D_tilde_0fill), type = "b", main = "Singular values of the observed mixture")
```

`rrmc()` is best suited for data characterized by slowly decaying singular values, indicative of complex underlying patterns.
Most EH data can be described this way. `root_pcp()` is best for data characterized by rapidly decaying singular values, indicative of very well-defined latent patterns.
For a simple example like the above, both PCP models are perfectly suitable. We will use `rrmc()`.

To estimate the low-rank and sparse matrices, `rrmc()` needs to be given a maximum rank `r` and regularization parameter `eta`. To find these, we will conduct a brief
grid search using the `grid_search_cv()` function, searching all models up to from rank 1 through 5, and examining all values of `eta` near the default value calculated via `get_pcp_defaults()`:

```{r grid search}
eta_0 <- get_pcp_defaults(D_tilde)$eta
cat(eta_0)
etas <- data.frame("eta" = sort(c(0.1 * eta_0, eta_0 * seq(1, 10, 2))))
gs <- grid_search_cv(D_tilde, pcp_fn = rrmc, grid = etas, r = 5)
top3_params <- dplyr::slice_head(dplyr::arrange(gs$summary_stats, rel_err), n = 3)
r_star <- top3_params$r[1]
eta_star <- round(top3_params$eta[1], 3)
top3_params
```

The grid search correctly identified the rank `r r_star` solution as the best (lowest relative error rate)! The corresponding `eta` = `r eta_star`.
Now we can run our PCP model:

```{r rrmc}
pcp_model <- rrmc(D_tilde, r = r_star, eta = eta_star)
```

As a baseline of comparison, let's see how well PCA can recover `L_0`.
We can give PCA an advantage by sharing PCP's discovery that the solution should be rank `r r_star`:

```{r pca}
L_pca <- proj_rank_r(D_tilde_0fill, r = r_star)
```

Now let's look at PCP's estimate of the sparse matrix and fix any values that are "practically" zero using the `hard_threshold()` function.

```{r sparse}
hist(pcp_model$S)
pcp_model$S <- hard_threshold(pcp_model$S, thresh = 0.2)
```

Finally, let's see how we did in recovering `L_0` and `S_0`:

```{r performance metrics}
data.frame(
  "Obs_rel_err" = norm(L_0 - D_tilde_0fill, "F") / norm(L_0, "F"),
  "PCA_L_rel_err" = norm(L_0 - L_pca, "F") / norm(L_0, "F"),
  "PCP_L_rel_err" = norm(L_0 - pcp_model$L, "F") / norm(L_0, "F"),
  "PCP_S_rel_err" = norm(S_0 - pcp_model$S, "F") / norm(S_0, "F"),
  "PCP_L_rank" = matrix_rank(pcp_model$L)
)
```

PCP outperformed PCA by quite a bit! We successfully recovered the chemical exposure patterns in `L_0` and extreme exposure events in `S_0` with only small relative errors in each.
We can now pair our estimated `L` matrix with any matrix factorization method of our choice (e.g. factor analysis or NMF) to extract the latent chemical exposure patterns.
These patterns, along with the isolated outlying exposure events in `S`, can then be incorporated with any outcomes of interest in downstream epidemiological analyses.

The `pcpr` package also ships with a real world air pollution dataset `queens` consisting of PM2.5 measurements (in µg/m^3) of
26 PM2.5 chemical species, recorded every three to six days from 04/04/2001 through 12/30/2021 courtesy of an EPA AQS monitoring site in Queens, NY:

```{r queens}
queens
```

You can follow along a real source apportionment analysis using the `queens` data in the source apportionment vignette.

## References
Gibson, Elizabeth A., Junhui Zhang, Jingkai Yan, Lawrence Chillrud, Jaime Benavides, Yanelli Nunez, Julie B. Herbstman, Jeff Goldsmith, John Wright, and Marianthi-Anna Kioumourtzoglou. "Principal component pursuit for pattern identification in environmental mixtures." Environmental Health Perspectives 130, no. 11 (2022): 117008. [available [here](https://doi.org/10.1289/EHP10479)]

Zhang, Junhui, Jingkai Yan, and John Wright. "Square root principal component pursuit: tuning-free noisy robust matrix recovery." Advances in Neural Information Processing Systems 34 (2021): 29464-29475. [available [here](https://proceedings.neurips.cc/paper/2021/hash/f65854da4622c1f1ad4ffeb361d7703c-Abstract.html)]

Cherapanamjeri, Yeshwanth, Kartik Gupta, and Prateek Jain. "Nearly optimal robust matrix completion." International Conference on Machine Learning. PMLR, 2017. [available [here](https://proceedings.mlr.press/v70/cherapanamjeri17a.html)]