---
title: "Air pollution source apportionment with PCP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Air pollution source apportionment with PCP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette, we will see how to leverage the functions in `pcpr` to conduct
an example air pollution source apportionment analysis. We will use PCP to apportion
speciated PM$_{2.5}$ to its sources using the `queens` dataset that comes with the
`pcpr` R package. The `queens` dataset consists of real chemical concentrations
(in µg/m$^3$) of 26 species of PM$_{2.5}$ measured every three to six days from
04/04/2001 through 12/30/2021 by an EPA AQS air monitor located in Queens, New York City.

Let's load and attach all the packages we will need for our analysis into the current R session.
We will also define a helper function `plot_matrix()` to visualize our data as a heatmap.

```{r setup}
library(pcpr)
suppressPackageStartupMessages({
  library(dplyr) # for manipulation of queens
  library(GGally) # for the ggcorr fn
  library(ggplot2) # for plotting
  library(lubridate) # for manipulating dates
  library(magrittr) # for the pipe %>%
  library(tidyr) # for pivoting queens for better plotting
})

plot_matrix <- function(D, ..., lp = "none", title = NULL) {
  D <- t(D)
  if (is.null(colnames(D))) colnames(D) <- paste0("C", 1:ncol(D))
  data.frame(D) %>%
    dplyr::mutate(x = paste0("R", 1:nrow(.))) %>%
    tidyr::pivot_longer(
      tidyselect::all_of(colnames(D)),
      names_to = "y",
      values_to = "value"
    ) %>%
    dplyr::mutate(
      x = factor(x, levels = unique(x)),
      y = factor(y, levels = unique(y))
    ) %>%
    ggplot(aes(x = x, y = y, fill = value)) +
    geom_raster() +
    scale_y_discrete(limits = rev) +
    coord_equal() +
    scale_fill_viridis_c(na.value = "white", ...) +
    theme_minimal() +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      legend.position = lp,
      plot.margin = margin(0, 0, 0, 0),
      aspect.ratio = 1
    ) +
    ggtitle(title)
}
```

## Preprocessing

### Exploring the raw data

We start by examining the raw `queens` data:

```{r queens raw}
queens
```

Let's visualize our data as a heatmap using the `plot_matrix()` function we wrote above:

```{r plot queens}
plot_matrix(queens[, 2:ncol(queens)])
```

Clearly the scale of each chemical differs widely.
Let's now visualize each of the measured chemical species as timeseries to get a better sense
of what we're dealing with. Plotted in black are the raw observed
PM$_{2.5}$ measurements (in µg/m$^3$) over time (04/04/2001 - 12/30/2021), and
plotted in red are some (rough) lines of best fit: 

```{r trends}
queens %>%
  pivot_longer(
    colnames(queens)[-1],
    names_to = "chem", values_to = "concentration"
  ) %>%
  filter(!is.na(concentration)) %>%
  ggplot(aes(x = Date, y = concentration)) +
  geom_line() +
  geom_smooth(color = "red", formula = "y ~ x", method = "loess", span = 0.05) +
  facet_wrap(~chem, scales = "free_y") +
  labs(x = "Date", y = "Concentration (µg/m^3)") +
  theme_bw()
```

Some initial points to make note of, looking at the data:

* As we saw in the heatmap, the scale for each chemical's measured concentration differs widely (e.g., NH4 ranges from 0 to
  9 µg/m$^3$, while Ti ranges from 0 to 0.04 µg/m$^3$). We will have to correct for this in a moment
  when we preprocess our data.
* Some chemical species record negative measurements (e.g., Al, Ba, Cd). As explained in the
  EPA [AQS data documentation](https://aqs.epa.gov/aqsweb/documents/about_aqs_data.html#acceptable-values),
  this is due to idiosyncrasies in the instruments used to collect the measurements. We will zero out
  negative measurements during preprocessing.
* We can see strong seasonal trends in some species (e.g., NO3), as well as a reduction in
  concentration over time in others (e.g., Ni).
* Many species have prominent outliers, or extreme exposure events (e.g., Cr, Cu). Further,
  some species have extreme exposure events following seasonal trends, e.g., K.
* Elemental carbon (EC) and organic carbon (OC) are both missing measurements for the first
  8 years of the dataset, likely because the air monitors for those two species were not
  operational from 2001 - 2009. Handling such systematic missingness is out of scope for
  this tutorial.

Below we remove the all measurements made before 2015, yielding the `queens_small` dataset that we will use moving forward.
We make this decision because we have prior knowledge that in the years leading up to 2015, gradually, many Midwestern coal-fired power
plants closed, reducing the regional / secondary signal experienced in downwind Queens, NYC after 2015
([Hopke et al. (2024)](https://doi.org/10.1016/j.envpol.2024.123708)).
We are interested in looking at the trends in the data after this (soft) change.

```{r dates}
start_date <- "2015-01-01"
queens_small <- queens %>% filter(Date >= as.Date(start_date))
queens_small
```

Next, let's take a look at the correlation structure of our raw `queens_small` data:

```{r raw correlation}
ggcorr(
  queens_small[, 2:ncol(queens_small)],
  method = "pairwise.complete.obs",
  limits = FALSE, label = FALSE, size = 5
)
```

The noisy measurements and high dimensionality of our `queens_small` air pollution mixture matrix
gives rise to this relatively complex correlation matrix. No strong patterns jump out right away here.
We'd like to employ PCP in order to reduce the complexity of our data - handling noise and outliers -
for more robust downstream analysis. Keep this correlation matrix in mind, since after applying
PCP, we'll examine the correlation matrix of the recovered `L` matrix and compare.

### Data standardization

Before we are able to run PCP, we need to first preprocess our data for better numerical stability.
Despite making minimal _assumptions_ of input data, PCP converges _in practice_ much more quickly when
the data has been standardized in some way. The goal is to rescale the data so columns are of
comparable scale _without meaningfully altering the original distributions of the columns._
Let's examine the distributions as they are in the `queens_small` dataset:

```{r distributions}
queens_small %>%
  pivot_longer(
    colnames(queens_small)[-1],
    names_to = "chem", values_to = "concentration"
  ) %>%
  filter(!is.na(concentration)) %>%
  ggplot(aes(x = concentration)) +
  geom_histogram(bins = 50) +
  theme_bw() +
  facet_wrap(~chem, scales = "free")
```

We have many choices for how we'd like to preprocess our data, e.g., standardize the data, min-max
normalization, etc. Most of our data appears normal, log normal, or exponential. Here we choose to
scale (but not center) each PM$_2.5$ species in our dataset (we also choose to set all negative
measurements to 0 for better interpretability):

```{r preprocessing}
queens_scaled <- queens_small %>% select(-Date)
queens_scaled[queens_scaled < 0] <- 0

queens_scaled <- queens_scaled %>%
  scale(center = FALSE) %>%
  as_tibble()

queens_scaled$Date <- queens_small$Date

queens_scaled %>%
  pivot_longer(
    colnames(queens_small)[-1],
    names_to = "chem", values_to = "concentration"
  ) %>%
  filter(!is.na(concentration)) %>%
  ggplot(aes(x = concentration)) +
  geom_histogram(bins = 50) +
  theme_bw() +
  facet_wrap(~chem, scales = "free")
```

That's all there is to our preprocessing! Now we can turn our attention to PCP.

## Principal component pursuit
### Model selection

There are two PCP algorithms shipped with `pcpr`: the convex `root_pcp()`
[[Zhang et al. (2021)](https://proceedings.neurips.cc/paper/2021/hash/f65854da4622c1f1ad4ffeb361d7703c-Abstract.html)]
and non-convex `rrmc()` [[Cherapanamjeri et al. (2017)](https://proceedings.mlr.press/v70/cherapanamjeri17a.html)].

`rrmc()` is best suited for data characterized by slowly decaying singular values,
indicative of complex underlying patterns and a relatively large degree of noise.
Most EH data can be described this way. `root_pcp()` is best for data characterized
by rapidly decaying singular values, indicative of very well-defined latent patterns.

To figure out which model would be best for our data, let's inspect the singular values
of our observed mixture using the `sing()` method. While PCP can handle `NA` values,
`sing()` cannot, so we quickly impute missing values with the corresponding column mean
via the `impute_matrix()` function:

```{r sing}
D <- as.matrix(queens_scaled %>% select(-Date))
D_imputed <- impute_matrix(D, apply(D, 2, mean, na.rm = TRUE))
singular_values <- sing(D_imputed)
plot(singular_values, type = "b")
```

These singular values slowly decay, corroborating the relatively complex underlying low-rank structure
obscured by a high degree of noise we saw in the correlation matrix earlier.
As such, we will move forward with `rrmc()` as our PCP model.

### Grid search for parameter tuning

To estimate the low-rank and sparse matrices `L` and `S`, `rrmc()` needs to be given a maximum rank
to search up to `r` and regularization parameter `eta`. To determine the optimal values of `r` and `eta`,
we will conduct a brief grid search using the `grid_search_cv()` function.

Recall that `rrmc()` uses an incremental rank-based procedure in recovering `L` and `S`:
First, a rank-$1$ model $(L^{(1)}, S^{(1)})$ is estimated. The rank-$1$ model is then used
as an initialization point to construct a rank-$2$ model $(L^{(2)}, S^{(2)})$, and so on,
until the desired rank-r model $(L^{(r)}, S^{(r)})$ is recovered. All models from ranks $1$
through $r$ are returned by `rrmc()` in this way. As such, we can fix the rank `r` in our
gridsearch to be the maximum rank we would like to search up to. Here we've chosen $r = 8$,
since we do not expect more than 8 distinct sources to govern `queens` PM$_2.5$ data from
2015-2021 (based on prior studies suggesting ~5 sources would perhaps be more reasonable).

We also need to tune the `eta` parameter, ensuring we also cover the rough default value
obtained via the `get_pcp_defaults()`. The `eta` parameter can be thought of as a dial
controlling the interplay between the low-rank `L` and sparse `S` models. Larger values of
`eta` will place a greater emphasis on penalizing the non-zero entries in `S` over penalizing
the errors between the predicted and observed data (the dense noise $Z$).

Note that this time we do not need to impute the data with chemical means, since we would
like `rrmc()` to recover missing `NA` values using the low-rank structure present in the data.

```{r gs, eval=FALSE, echo=TRUE}
eta_0 <- get_pcp_defaults(D)$eta
# to get progress bar, could wrap this
# in a call to progressr::with_progress({ gs <- grid_search_cv(...) })
gs <- grid_search_cv(
  D,
  pcp_fn = rrmc,
  grid = data.frame(eta = 10^seq(-1, 2, 1) * round(eta_0, 3)),
  r = 8,
  parallel_strategy = "multisession",
  num_workers = 16,
  verbose = FALSE
)
r_star <- gs$summary_stats$r[1]
eta_star <- round(gs$summary_stats$eta[1], 3)
gs$summary_stats
```

```{r gs results, eval=TRUE, echo=FALSE}
gs <- readRDS("rds_files/applied-gs.rds")
r_star <- gs$summary_stats$r[1]
eta_star <- round(gs$summary_stats$eta[1], 3)
gs$summary_stats
```

The results from the search suggest a rank `r gs$summary_stats$r[1]` solution with an `eta` of
`r gs$summary_stats$eta[1]` is most optimal, with the lowest average relative recovery error on
the held out test sets (`r round(gs$summary_stats$rel_err[1] * 100, 1)`% relative error). This solution yields a
sparse matrix `S` with a sparsity of `r round(gs$summary_stats$S_sparsity[1] * 100, 1)`%, meaning
`r 100 - round(gs$summary_stats$S_sparsity[1] * 100, 1)`% of values in the data were flagged as extreme,
outlying exposure events. As such, we will use these parameters moving forward.

### Running PCP

Now we can run our PCP model!

```{r pcp}
pcp_model <- rrmc(D, r = r_star, eta = eta_star)
L <- pcp_model$L
S <- pcp_model$S
```

We can inspect the evolution of the objective function over the course of PCP's optimization:

```{r obj}
plot(pcp_model$objective, type = "l")
```

### The recovered L matrix

And the output `L` matrix:

```{r output L}
plot_matrix(L)
L_rank <- matrix_rank(L)
L_rank
```

Recall the original correlation matrix for `queens_small` was:

```{r orig corr}
ggcorr(
  queens_small[, 2:ncol(queens_small)],
  method = "pairwise.complete.obs",
  limits = FALSE, label = FALSE, size = 5
)
```

The new correlation matrix for the recovered `L` matrix is now:

```{r new corr}
L_df <- data.frame(L)
colnames(L_df) <- colnames(queens[, 2:ncol(queens)])
ggcorr(
  L_df, method = "pairwise.complete.obs",
  limits = FALSE, label = FALSE, size = 5
)
```

As we can see, PCP successfully reduced the dimensionality of the data. The underlying correlation structure
is now made much more explicit.

### The recovered S matrix

Let's briefly inspect PCP's estimate of the sparse matrix `S` and the associated sparsity via `sparsity()`:

```{r sparse}
sparsity(S)
hist(S)
plot_matrix(S)
```

We can examine which chemicals have the largest number of outlying exposure events by calculating
the sparsity of each column in `S`:

```{r sparse by col}
S_df <- data.frame(S)
colnames(S_df) <- colnames(queens_small[, 2:ncol(queens_small)])
chems_w_most_extreme_events <- sort(apply(S_df, 2, function(x) sparsity(as.matrix(x))))
chems_w_most_extreme_events
```

Here we can see ammonium (NH4), silicon (Si), and nitrate (NO3) have no extreme exposure events.
Arsenic (As), vanadium (V), and selenium (Se) have the greatest number of outlying events.
We can examine the extreme events for an individual chemical with:

```{r As sparse events}
S_df %>%
  select(As) %>%
  mutate(Date = queens_small$Date) %>%
  filter(As > 0)
```

Let's check out another example, this time examining the sparse events isolated for potassium (K), associated with biomass burning and fireworks.

```{r K sparse events}
S_df %>%
  select(K) %>%
  mutate(Date = queens_small$Date) %>%
  filter(K > 0)
```

PCP was able to pick up extreme potassium events around the July 4th holiday 
for each of the July 4th / 5th dates available in the dataset (recall that chemical
concentrations are recorded in `queens` every three to six days, skipping 
measurements on July 4/5 some years).

Other dimensionality reduction techniques
such as PCA can be sensitive to outliers, requiring researchers to manually remove
outlying events from their data. This introduces subjectivity into the data preprocessing
pipeline, as the researcher must to define a criteria for removal, in turn making downstream
results less replicable.

We're now ready to finish the source apportionment using the estimated low-rank `L` matrix.

## PCP and NMF for source apportionment

We will use Non-negative Matrix Factorization (NMF) to extract the patterns (loadings) and
corresponding scores encoded in the `L` matrix. See `NMF::nmf()` for details.

Since we determined `L` to be of rank `r L_rank` from PCP's grid search, we no longer need to
worry about what rank to use for our NMF step, nor do we need to manually remove outliers, since
PCP autonomously isolated those into the sparse matrix `S`.

### Non-negative matrix factorization
```{r NMF, eval=FALSE, echo=TRUE}
library(NMF)

nmf_mat <- L
nmf_mat[nmf_mat < 0] <- 0

res <- nmf(
  nmf_mat, rank = L_rank,
  method = "offset", nrun = 30,
  seed = 0, .opt = "vp16"
)

raw_scores <- basis(res)
raw_loadings <- coef(res)
```

```{r load NMF, eval=TRUE, echo=FALSE}
raw_scores <- readRDS("rds_files/applied-nmf-raw-scores-W.rds")
raw_loadings <- readRDS("rds_files/applied-nmf-raw-loadings-H.rds")
```

### PCP-NMF Loadings

Let's inspect the loadings obtained from our PCP-NMF model:

```{r loadings}
loadings <- data.frame(raw_loadings)
colnames(loadings) <- colnames(L_df)
loadings[["Pattern"]] <- paste("Pattern", 1:L_rank)

loadings %>%
  pivot_longer(-Pattern, names_to = "Chemical", values_to = "Loading") %>%
  ggplot(aes(x = Chemical, y = Loading)) +
  geom_point(size = 3) +
  geom_segment(aes(yend = 0, xend = Chemical), linewidth = 1.5) +
  facet_wrap(~Pattern, scales = "free_x") +
  theme_bw() +
  theme(
    strip.text.x = element_text(size = 14),
    title = element_text(size = 18),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    strip.background = element_rect(fill = "white"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  ) +
  geom_hline(yintercept = 0, size = 0.2) +
  ggtitle("PCP-NMF Chemical Loadings")
```

* Pattern 1 appears to be linked to traffic and road-related sources,
  characterized by elements such as barium, calcium, and lead, indicating emissions
  from brake wear, tire wear, and construction activities
* Pattern 2 looks like crustal dust, marked by elements like aluminum, calcium, silicon,
  titanium, and iron, suggesting resuspension of natural materials from soil and rock.
* Pattern 3 is clearly salt, including sodium chloride and magnesium
  chloride.
* Pattern 4 represents a mix of regional/secondary sources and traffic/tailpipe emissions,
  characterized by sulfur, ammonium, nitrate, organic carbon, and zinc. This pattern
  lacks a clear regional signal and suggests contributions from both
  regional/secondary sources and tailpipe emissions. 

### Variance Explained

Let's calculate the variance explained by each of the patterns using the pattern scores next.
We will then reorder the patterns to appear according to their proportion of variance explained.

```{r var exp}
nmf_scores <- data.frame(raw_scores)
colnames(nmf_scores) <- c(
  "Traffic", "Crustal Dust", "Salt", "Regional/Secondary & Tailpipe Emissions"
)
days_of_the_week <- c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat")
nmf_sources <- nmf_scores %>%
  mutate(
    Date = queens_scaled$Date,
    DayofWeek = as.character(wday(Date, label = TRUE)),
    DayofWeek = factor(DayofWeek, levels = days_of_the_week),
  ) %>%
  pivot_longer(
    -c(Date, DayofWeek),
    names_to = "Pattern",
    values_to = "Score"
  )

var_explained <- nmf_sources %>%
  group_by(Pattern) %>%
  summarize(patsum = sum(Score)) %>%
  mutate(VarianceExplained = round(100 * patsum / sum(patsum), 1)) %>%
  select(-patsum) %>%
  arrange(desc(VarianceExplained)) %>%
  mutate(PatName = paste(Pattern, paste0("(", VarianceExplained, "% Var Exp)")))

var_explained %>% select(-PatName)

nmf_sources <- nmf_sources %>%
  mutate(
    Pattern = factor(
      Pattern,
      levels = as.character(var_explained$Pattern),
      labels = as.character(var_explained$PatName)
    )
  )
```

### Loadings replotted

We can now replot the loadings with the proper pattern names:

```{r rename patterns}
loadings %>%
  mutate(Pattern = factor(
    colnames(nmf_scores),
    levels = as.character(var_explained$Pattern),
    labels = as.character(var_explained$PatName)
  )) %>%
  pivot_longer(-Pattern, names_to = "Chemical", values_to = "Loading") %>%
  ggplot(aes(x = Chemical, y = Loading, color = Pattern)) +
  scale_color_brewer(palette = "Set1") +
  geom_point(size = 3) +
  geom_segment(aes(yend = 0, xend = Chemical), linewidth = 1.5) +
  facet_wrap(~Pattern, scales = "free_x") +
  theme_bw() +
  theme(
    strip.text.x = element_text(size = 14),
    title = element_text(size = 18),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    strip.background = element_rect(fill = "white"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  ) +
  geom_hline(yintercept = 0, size = 0.2) +
  ggtitle("PCP-NMF Loadings")
```

### PCP-NMF Scores

We can examine the scores over time:

```{r NMF scores}
yr_num <- length(unique(year(nmf_sources$Date)))

ggplot(nmf_sources, aes(Date, Score, color = Pattern)) +
  scale_color_brewer(palette = "Set1") +
  geom_smooth(method = "loess", span = 0.05) +
  facet_wrap(~Pattern, scales = "free", ncol = 1) +
  xlab("") +
  ylab("Pattern Score") +
  theme_classic() +
  theme(
    legend.position = "none",
    title = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    strip.text = element_text(size = 14),
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 11)
  ) +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  ggtitle("PCP-NMF Scores over Time")
```

The scores across the seasons (where "DJF" = Winter, "MAM" = Spring,
"JJA" = Summer, and "SON" = Autumn)

```{r seasonal trends}
nmf_sources %>%
  mutate(Season = metR::season(Date)) %>%
  group_by(Season, Pattern) %>%
  summarise(
    mean_score = mean(Score),
    sd_score = sd(Score),
    .groups = "keep"
  ) %>%
  ggplot(aes(Season, mean_score, color = Pattern)) +
  scale_color_brewer(palette = "Set1") +
  geom_point(size = 6) +
  geom_errorbar(
    aes(ymin = mean_score - sd_score, ymax = mean_score + sd_score),
    width = 0.25, linewidth = 1.5
  ) +
  xlab("") +
  facet_wrap(~Pattern, scales = "free_x", ncol = 1) +
  ylab("Factor Score") +
  theme_bw() +
  theme(
    title = element_text(size = 18),
    legend.position = "none",
    axis.title.y = element_text(size = 18),
    axis.text.x.bottom = element_text(size = 13, color = "black"),
    strip.text = element_text(size = 16),
    axis.text.y.left = element_text(size = 12)
  ) +
  ggtitle("Mean (std dev) Pattern Scores by Season")

```

Scores across days of the week:

```{r weekly trends}
nmf_sources %>%
  group_by(Pattern, DayofWeek) %>%
  summarize(
    mean_score = mean(Score),
    sd_score = sd(Score),
    .groups = "keep"
  ) %>%
  ggplot(aes(DayofWeek, mean_score, color = Pattern)) +
  scale_color_brewer(palette = "Set1") +
  geom_point(size = 6) +
  geom_errorbar(
    aes(ymin = mean_score - sd_score, ymax = mean_score + sd_score),
    width = 0.25, linewidth = 1.5
  ) +
  xlab("") +
  facet_wrap(~Pattern, scales = "free_x", ncol = 1) +
  ylab("Factor Score") +
  theme_bw() +
  theme(
    title = element_text(size = 18),
    legend.position = "none",
    axis.title.y = element_text(size = 18),
    axis.text.x.bottom = element_text(size = 13, color = "black"),
    strip.text = element_text(size = 16),
    axis.text.y.left = element_text(size = 12)
  ) +
  ggtitle("Mean (std dev) Pattern Scores across Days of the Week")
```

Scores for weekday vs. weekend:

```{r day type}
nmf_sources %>%
  mutate(
    DayType = if_else(DayofWeek %in% c('Mon', 'Tue', 'Wed', 'Thu', 'Fri'), 'Weekday', 'Weekend'),
    DayType = factor(DayType)
  ) %>%
  group_by(Pattern, DayType) %>%
  summarize(
    mean_score = mean(Score),
    sd_score = sd(Score),
    .groups = "keep"
  ) %>%
  ggplot(aes(DayType, mean_score, color = Pattern)) +
  scale_color_brewer(palette = "Set1") +
  geom_point(size = 6) +
  geom_errorbar(
    aes(ymin = mean_score - sd_score, ymax = mean_score + sd_score),
    width = 0.25, linewidth = 1.5
  ) +
  xlab("") +
  facet_wrap(~Pattern, scales = "free_x", ncol = 1) +
  ylab("Factor Score") +
  theme_bw() +
  theme(
    title = element_text(size = 18),
    legend.position = "none",
    axis.title.y = element_text(size = 18),
    axis.text.x.bottom = element_text(size = 13, color = "black"),
    strip.text = element_text(size = 16),
    axis.text.y.left = element_text(size = 12)
  ) +
  ggtitle("Mean (std dev) Pattern Scores Weekend vs. Weekday")
```

Pattern correlations:

```{r nmf corr}
ggcorr(
  nmf_scores, method = c("pairwise.complete.obs", "pearson"),
  limits = FALSE, label = TRUE, label_size = 8,
  hjust = 0.8, size = 8
)
```

## Conclusion

We paired PCP with NMF for the source apportionment of the `queens` dataset from
2015 - 2021. We identified four patterns 

The first identified pattern, accounting for 31% of the commonly shared variance
of the low-rank matrix, is linked to traffic and road-related sources,
characterized by elements such as barium, calcium, and lead, indicating emissions
from brake wear, tire wear, and construction activities. The second identified
pattern, representing 21% of the variance, is associated with crustal dust,
marked by elements like aluminum, calcium, silicon, titanium, and iron, suggesting
resuspension of natural materials from soil and rock. The third pattern, comprising
14% of the variance, is identified as salt, including sodium chloride and magnesium
chloride. This pattern peaks in early spring, likely due to the accumulation of
road salt over winter. The fourth pattern, accounting for 35% of the variance,
represents a mix of regional/secondary sources and traffic/tailpipe emissions,
characterized by sulfur, ammonium, nitrate, organic carbon, and zinc. This pattern
lacks a clear regional signal and suggests contributions from both
regional/secondary sources and tailpipe emissions. 
