---
title: "Quick start: applying PCP to a simulated environmental mixture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick start: applying PCP to a simulated environmental mixture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The R package `pcpr` implements Principal Component Pursuit (PCP), a robust
dimensionality reduction technique, for pattern recognition tailored to
environmental health data. The statistical methodology and computational details
are provided in [Gibson et al. (2022)](https://doi.org/10.1289/EHP10479).

In this vignette, we will see how to leverage the functions in `pcpr` to:

1. Simulate and explore a simple environmental mixture dataset;
2. Determine the best PCP algorithm to employ with the simulated data;
3. Tune PCP's parameters using a cross-validated grid search; and
4. Evaluate PCP's performance against the simulated ground truth mixture model.

## Preliminaries

### PCP modeling overview

PCP algorithms model an observed exposure matrix $D$ as the sum of three underlying
ground-truth matrices:

$$\underset{\text{mixture}}{D_{n \times p}} = \underset{\text{low-rank}}{L_0} + \underset{\text{sparse}}{S_0} + \underset{\text{noise}}{Z_0}$$

a low-rank matrix $L_0$ encoding consistent patterns of exposure, a sparse
matrix $S_0$ isolating unique or outlying exposure events (that cannot be
explained by the consistent exposure patterns), and dense noise $Z_0$. All of
these matrices are of dimension $n \times p$, where $n$ is the number of
observations (e.g. study participants or measurement dates) and $p$ is the
number of exposures (e.g. chemical and/or non-chemical stressors). Beyond this
mixtures model, the main assumption made by PCP is that
$Z_0 \sim N(\mu, \sigma^2)$ consists of i.i.d. Gaussian noise
corrupting each entry of the overall exposure matrix $D$.

The models in `pcpr` seek to decompose an observed data matrix `D` into estimated
low-rank and sparse components `L` and `S` for use in downstream environmental health
analyses.

### Extensions for environmental health data

The functions in `pcpr` are outfitted with three environmental health
(EH)-specific extensions, making `pcpr` particularly powerful for EH research:

1. **Missing value functionality:** PCP is able to recover `NA` values in
   the observed mixture matrix, often outperforming traditional imputation techniques.
2. **Leveraging potential limit of detection (LOD) information:** When equipped with LOD
  information, PCP treats any estimations of values known to be below the LOD as equally
  valid if their approximations fall between 0 and the LOD. PCP with LOD data often
  outperforms PCA imputed with $\frac{LOD}{\sqrt{2}}$.
3. **Non-negativity constraint on the estimated `L` matrix:** If desired, PCP can enforce
  values in the estimated low-rank matrix `L` to be $\geq 0$, better modeling real world
  mixtures data.

## Simulated environmental mixture example  

We start by loading and attaching the `pcpr` package to our current R session.

```{r setup}
library(pcpr)
```

### Simulating data

The `sim_data()` function lets users generate simple mixtures models for
quick experimentation. Let's use it to simulate a noisy environmental mixture
comprised of 100 observations of 10 chemicals, with three underlying chemical
exposure patterns (or a rank $r = 3$), extreme outlying exposure events along
the diagonal of the matrix, and dense Gaussian noise corrupting all the
measurements in the matrix:

```{r sim data}
library(pcpr)

data <- sim_data(
  n = 100, p = 10, r = 3,
  sparse_nonzero_idxs = seq(1, 1000, 101),
  sigma = 0.05
)
D <- data$D
L_0 <- data$L
S_0 <- data$S
Z_0 <- data$Z
```

The rank of a matrix is the number of linearly independent columns or rows in the matrix
governing the structure of the data. It can intuitively be thought of as the
number of inherent latent patterns in the data. The `matrix_rank()` function estimates
the rank of a given matrix by counting the number of nonzero singular values governing
that matrix:

```{r matrix rank}
matrix_rank(L_0)
matrix_rank(D)
```

Here we can see `L_0` has 3 underlying patterns. This is obscured in the observed, full-rank
mixture matrix `D`, since the `S_0` and `Z` components are corrupting the underlying structure
provided by `L_0`.

Because our mixtures data is often incomplete in practice, let's further corrupt a random 5% of
the values as missing `NA` with the `corrupt_mat_randomly()` function:

```{r corrupt mat randomly}
corrupted_data <- corrupt_mat_randomly(D, perc = 0.05)
D_tilde <- corrupted_data$D_tilde
D_tilde_imputed <- D_tilde
D_tilde_imputed[is.na(D_tilde)] <- 0
```

The `D_tilde` matrix represents our observed, messy mixtures model, suffering from
incomplete `NA` observations. In a moment we will run our PCP model on `D_tilde`
directly. The `D_tilde_imputed` matrix imputes `NA` values with `0`, allowing us to
call functions that do not support missing `NA` values.

### PCP model selection

There are two PCP algorithms shipped with `pcpr`: the convex `root_pcp()`
[[Zhang et al. (2021)](https://proceedings.neurips.cc/paper/2021/hash/f65854da4622c1f1ad4ffeb361d7703c-Abstract.html)]
and non-convex `rrmc()` [[Cherapanamjeri et al. (2017)](https://proceedings.mlr.press/v70/cherapanamjeri17a.html)].
To figure out which model would be best for our data, let's inspect the singular
values of our observed mixture using the `sing()` method:

```{r sing}
singular_values <- sing(D_tilde_imputed)
plot(singular_values, type = "b")
```

`rrmc()` is best suited for data characterized by slowly decaying singular values,
indicative of complex underlying patterns and a relatively large degree of noise.
Most EH data can be described this way. `root_pcp()` is best for data characterized
by rapidly decaying singular values, indicative of very well-defined latent patterns.

For a simple example like the above, both PCP models are perfectly suitable. 
We will use `rrmc()`, as this is the model environmental health researchers will
likely employ most frequently.

### Tuning PCP parameters via grid search

To estimate the low-rank and sparse matrices, `rrmc()` needs to be given a maximum rank
`r` and regularization parameter `eta`. To determine the optimal values of `r` and `eta`,
we will conduct a brief grid search using the `grid_search_cv()` function. In our grid
search below, we will examine all models from ranks 1 through 5 and all values of `eta`
near the default, calculated with `get_pcp_defaults()`.

The `rrmc()` approach to PCP uses an iterative rank-based procedure to recover `L` and
`S`, meaning it first constructs a rank `1` model and iteratively builds up to the
specified rank `r` solution. As such, for our grid search with `grid_search_cv()`, we
pass `etas` as the `grid` argument we are searching and we send `r` = 5 as a constant
parameter common to all models in the search. Since `length(etas) = 6` and `r` = 5, we
are grid searching through 30 different PCP models. The `num_runs` argument in
`grid_search_cv()` determines how many (random) tests should be performed for each unique
model setting. By default, `num_runs = 100`, so our grid search tunes `r` and `eta` by
measuring the performance of 300 different PCP models.

```{r grid search}
eta_0 <- get_pcp_defaults(D_tilde)$eta
cat(paste("Default eta value: ", eta_0))
etas <- data.frame("eta" = sort(c(0.1 * eta_0, eta_0 * seq(1, 10, 2))))
gs <- grid_search_cv(D_tilde, pcp_fn = rrmc, grid = etas, r = 5)
r_star <- gs$summary_stats$r[1]
eta_star <- round(gs$summary_stats$eta[1], 3)
head(gs$summary_stats)
```

Inspecting the `summary_stats` table from the output grid search provides the mean
aggregated statistics for each of the 30 distinct parameter settings we tested.
The grid search correctly identified the rank `r r_star` solution as the best
(lowest relative error rate). The corresponding `eta` = `r eta_star`.

### Running PCP

Now we can run our PCP model:

```{r rrmc}
pcp_model <- rrmc(D_tilde, r = r_star, eta = eta_star)
```

Let's briefly inspect PCP's estimate of the sparse matrix `S`, and fix any values that
are "practically" zero using the `hard_threshold()` function. The histogram below shows
a majority of the entries in `S` are between -0.2 and 0.2, so we will call those values
"practically" zero, and the rest true outlying exposure events.

```{r sparse}
hist(pcp_model$S)
pcp_model$S <- hard_threshold(pcp_model$S, thresh = 0.2)
```

### Benchmarking with PCA

Before evaluating our PCP model, let's see how well a more traditional method such as
Principal Compoenents Analysis (PCA) can recover `L_0`, to provide a benchmark for
comparison.

The `proj_rank_r()` function (project matrix to rank `r`) approximates an input matrix
as low-rank using a rank-`r` truncated SVD, the same way PCA approximates a low-rank
matrix. Normally, a researcher would need to determine `r` subjectively. We will give
PCA an advantage by sharing PCP's discovery from the above grid search that the solution
should be of rank `r r_star`:

```{r pca}
L_pca <- proj_rank_r(D_tilde_imputed, r = r_star)
```

### Evaluating PCP against the ground truth

Finally, let's see how we did in recovering `L_0` and `S_0`. We will examine the
relative error between our model's estimates and the simulated ground truth matrices.
We use the Frobenius norm to calculate the relative errors between the matrices:

```{r performance metrics}
data.frame(
  "Obs_rel_err" = norm(L_0 - D_tilde_imputed, "F") / norm(L_0, "F"),
  "PCA_L_rel_err" = norm(L_0 - L_pca, "F") / norm(L_0, "F"),
  "PCP_L_rel_err" = norm(L_0 - pcp_model$L, "F") / norm(L_0, "F"),
  "PCP_S_rel_err" = norm(S_0 - pcp_model$S, "F") / norm(S_0, "F"),
  "PCP_L_rank" = matrix_rank(pcp_model$L)
)
```

PCP outperformed PCA by quite a bit! PCP's relative recovery error on the `L_0` matrix
stood at only `r round(norm(L_0 - pcp_model$L, "F") / norm(L_0, "F") * 100, 2)`%,
compared to an observed relative error of
`r round(norm(L_0 - D_tilde_imputed, "F") / norm(L_0, "F") * 100, 2)`% and PCA's relative
error of `r round(norm(L_0 - L_pca, "F") / norm(L_0, "F") * 100, 2)`%.
PCP's sparse matrix estimate was only off from the ground truth `S_0` by
`r round(norm(S_0 - pcp_model$S, "F") / norm(S_0, "F") * 100, 2)`%.

We can now pair our estimated `L` matrix with any matrix factorization method of our
choice (e.g. factor analysis or non-negative matrix factorization, NMF) to extract
the latent chemical exposure patterns. These patterns, along with the isolated outlying
exposure events in `S`, can then be incorporated with any outcomes of interest in
downstream epidemiological analyses.